---
title: "treetools part one"
author: "Alexander Shah"
format: html
---

## Introduction

Most of the work in tree-ring research happens before any analysis: reading files, cleaning formats, reshaping data. treetools handles that part. This vignette walks through reading raw RWL data into R, turning it into something usable, saving it to CSV, and then using this clean data to detect cross dating errors and create helpful visualizations of the data.

Every function in the package passes the same data frame around: three columns (sample_id, year, and ring_width in mm). Output from one function drops straight into the next.

---

## Setup

```{r}
#| include: false
install.packages("rlang", repos = "https://cloud.r-project.org")
install.packages("dplR", repos = "https://cloud.r-project.org")
install.packages("tidyr", repos = "https://cloud.r-project.org")
install.packages("dplyr", repos = "https://cloud.r-project.org")
library(dplR)
library(tidyr)
library(dplyr)
```

---

## Step 1: Load the Data with `read_rwl()`

RWL is the standard file format for tree-ring measurements. read_rwl() reads one and returns a data frame (samples as columns, years as rows). For this walkthrough we use gp.rwl, a dataset bundled with dplR, instead of a file on disk.

```{r}
#' @export
read_rwl <- function(path) {
  read.rwl(path)
}
```

```{r}
# Load the sample dataset
data(gp.rwl, package = "dplR")

# Take a look at the raw format
gp.rwl[150:155, 1:5]
```

The raw format is wide: one column per sample, years as row names. Filtering or plotting it directly is a pain, which is why it needs reshaping.

---

## Step 2: Tidy the Data with `tidy_rwl()`

tidy_rwl() takes that wide data frame and pivots it long: one row per measurement. From there you can filter by sample, group by year, or pipe it into ggplot without any extra work.

```{r}
#' @export
tidy_rwl <- function(rwl_df) {
  # The row names of the RWL data frame are the years,
  # pull them out as a column
  rwl_df$year <- as.integer(rownames(rwl_df))
  # Pivot from wide to long format
  tidy <- pivot_longer(rwl_df, cols = -year, names_to = "sample_id", values_to = "ring_width")
  # Convert back to a plain data frame
  tidy <- as.data.frame(tidy)
  # Drop rows where ring width is NA
  tidy <- tidy[!is.na(tidy$ring_width), ]
  tidy
}
```

```{r}
# Convert to tidy format
tidy_data <- tidy_rwl(gp.rwl)

# Check the output
head(tidy_data)
```

One row per measurement, three columns. Everything else in the package takes this as input.

---

## Step 3: Export the Data with `export_rwl()`

`export_rwl()` saves the tidy data frame to a CSV. Useful for sharing with collaborators or loading into Excel or Python.

```{r}
#' @export
export_rwl <- function(data, path) {
  write.csv(data, path, row.names = FALSE)
  invisible(path)
}
```

```{r}
# Save the tidy data to a CSV file
export_rwl(tidy_data, "treetools_output.csv")

# Confirm it was saved by reading it back in
saved <- read.csv("treetools_output.csv")
head(saved)
```

Same three columns written to a plain CSV. Opens in Excel, Python, anything.

---
## Step 4: Detect Cross Dating Errors with detect_crossdate_errors()

detect_crossdate_errors() takes the now tidy data and analyzes to look for cross dating errors using certain criteria including checking for negative widths, checking for missing rings, looking for zeros, and checking if there is a statistical outlier by looking for standard deviations greater than three. It then creates a summary report with the information of the tree and tree ring and it's error category.

```{r}
detect_crossdate_errors <- function(data) {
  
  # Check for negative widths
  physical_errors <- data %>%
    dplyr::filter(ring_width < 0) %>%
    dplyr::mutate(issue = "Negative growth detected")

  # Check for missing rings/zeros
  missing_rings <- data %>%
    dplyr::filter(ring_width == 0) %>%
    dplyr::mutate(issue = "Potential missing ring (zero width)")

  # Statistical outlier check (width > 3 SD)
  outliers <- data %>%
    dplyr::group_by(sample_id) %>%
    dplyr::mutate(z_score = (ring_width - mean(ring_width)) / sd(ring_width)) %>%
    dplyr::filter(abs(z_score) > 3) %>%
    dplyr::mutate(issue = "Statistical outlier (Z > 3)")

  # Combine all checks into one summary report
  flagged_summary <- dplyr::bind_rows(physical_errors, missing_rings, outliers)
  return(flagged_summary)}
```

This allows dendrochronologists to quickly check if there were errors when a machine or someone else was crossdating instead of tediously going individualy through hundreds or thousands of rings. 

---
## Step 5: Visualize Data with `plot_chronology()` and `map_sites()`

With clean, validated data in hand, treetools provides two functions for visual inspection: one for the temporal view and one for the spatial view.

---
  
  ### `plot_chronology()`
  
  `plot_chronology()` takes the tidy data frame and plots ring width over time for every sample. Each sample gets its own coloured line, making it easy to spot diverging growth patterns or shared climate signals across cores.

```{r}
#' @export
plot_chronology <- function(data, title = "Tree-Ring Chronology") {
  ggplot(data, aes(x = year, y = ring_width, colour = sample_id, group = sample_id)) +
    geom_line(alpha = 0.6) +
    labs(title = title, x = "Year", y = "Ring Width (mm)", colour = "Sample") +
    theme_bw()
}
```

```{r}
# Plot all samples from the tidy data
plot_chronology(tidy_data)
```

Each line is one core. Dips and peaks that appear across multiple samples at the same year point to shared environmental events — drought years, cold summers, or disturbance events.

---
  
  ### `map_sites()`
  
  `map_sites()` plots collection site coordinates on a world basemap. It takes a data frame with three columns — `site_id`, `longitude`, and `latitude` — and auto-zooms to fit all sites.

```{r}
#' @export
map_sites <- function(sites) {
  world <- map_data("world")
  ggplot() +
    geom_polygon(data = world, aes(x = long, y = lat, group = group),
                 fill = "#EDEDE9", colour = "#AAAAAA") +
    geom_point(data = sites, aes(x = longitude, y = latitude),
               colour = "#2C7BB6", size = 3) +
    geom_text(data = sites, aes(x = longitude, y = latitude, label = site_id),
              vjust = -0.8, size = 3) +
    labs(x = "Longitude", y = "Latitude") +
    theme_bw()
}
```

```{r}
# Build a sites data frame and map it
sites_demo <- data.frame(
  site_id   = c("CAM", "PYR", "MET"),
  longitude = c(-118.2, -119.8, -116.5),
  latitude  = c(37.5, 38.1, 36.9)
)

map_sites(sites_demo)
```

Geographic context matters in dendrochronology — sites at different elevations or aspects can show very different growth responses to the same climate event. `map_sites()` makes it easy to see where your samples came from at a glance.
---
## Summary
Tree tools allows dendrochronologists and those studying tree ring and climate research to streamline the data wrangling process in research to get to analysis faster. This is done by making rwl files tidy and transferrable to csv, detecting cross dating errors, and creating visualizations of tree ring chronology and mapping tree ring sample collections.

