---
title: "treetools part one"
author: "Alexander Shah"
format: html
---

## Introduction

Most of the work in tree-ring research happens before any analysis: reading files, cleaning formats, reshaping data. treetools handles that part. This vignette walks through reading raw RWL data into R, turning it into something usable, and saving it to CSV.

Every function in the package passes the same data frame around: three columns (sample_id, year, and ring_width in mm). Output from one function drops straight into the next.

---

## Setup

```{r}
#| include: false
install.packages("rlang", repos = "https://cloud.r-project.org")
install.packages("dplR", repos = "https://cloud.r-project.org")
install.packages("tidyr", repos = "https://cloud.r-project.org")
install.packages("dplyr", repos = "https://cloud.r-project.org")
library(dplR)
library(tidyr)
library(dplyr)
```

---

## Step 1: Load the Data with `read_rwl()`

RWL is the standard file format for tree-ring measurements. read_rwl() reads one and returns a data frame (samples as columns, years as rows). For this walkthrough we use gp.rwl, a dataset bundled with dplR, instead of a file on disk.

```{r}
#' @export
read_rwl <- function(path) {
  read.rwl(path)
}
```

```{r}
# Load the sample dataset
data(gp.rwl, package = "dplR")

# Take a look at the raw format
gp.rwl[150:155, 1:5]
```

The raw format is wide: one column per sample, years as row names. Filtering or plotting it directly is a pain, which is why it needs reshaping.

---

## Step 2: Tidy the Data with `tidy_rwl()`

tidy_rwl() takes that wide data frame and pivots it long: one row per measurement. From there you can filter by sample, group by year, or pipe it into ggplot without any extra work.

```{r}
#' @export
tidy_rwl <- function(rwl_df) {
  # The row names of the RWL data frame are the years,
  # pull them out as a column
  rwl_df$year <- as.integer(rownames(rwl_df))
  # Pivot from wide to long format
  tidy <- pivot_longer(rwl_df, cols = -year, names_to = "sample_id", values_to = "ring_width")
  # Convert back to a plain data frame
  tidy <- as.data.frame(tidy)
  # Drop rows where ring width is NA
  tidy <- tidy[!is.na(tidy$ring_width), ]
  tidy
}
```

```{r}
# Convert to tidy format
tidy_data <- tidy_rwl(gp.rwl)

# Check the output
head(tidy_data)
```

One row per measurement, three columns. That's it. Everything else in the package takes this as input.

---

## Step 3: Export the Data with `export_rwl()`

`export_rwl()` saves the tidy data frame to a CSV. Useful for sharing with collaborators or loading into Excel or Python.

```{r}
#' @export
export_rwl <- function(data, path) {
  write.csv(data, path, row.names = FALSE)
  invisible(path)
}
```

```{r}
# Save the tidy data to a CSV file
export_rwl(tidy_data, "treetools_output.csv")

# Confirm it was saved by reading it back in
saved <- read.csv("treetools_output.csv")
head(saved)
```

Same three columns written to a plain CSV. Opens in Excel, Python, anything.

---

## Summary

Read it in, tidy it, save it. The data frame you end up with is what the error detection and visualization functions expect.
